Python as a machine learning language: experiences from the scikit-learn project

### 2013-06-16 paper.tex
spellings

From its core, the library has been designed to tie in with the set of numeric
and scientific packages centered around the NumPy and SciPy libraries. NumPy
\citep{vanderwalt2011} augments Python with a numeric array datatype and proper
array computing primitives while SciPy \citep{varoquaux2013scipy} extends it
further by making common numerical operations available, either by implementing
those in Python/NumPy or by wrapping existing C/C{}\verb!++!/Fortran
implementations. In a similar vein, \sklearn extends Python with a set of
composable machine learning operations.

@@

\begin{description}
  \item[Consistency.]
       All objects (basic or composed) share a consistent interface composed of
       a limited set of methods. This interface is documented in a consistent
       manner for all objects.
  \item[Inspection.]
       Constructor parameters and parameter values determined by learning
       algorithms are stored and exposed as public attributes.
  \item[Non-proliferation of classes.]
       Learning algorithms are the only objects to be of custom classes.
       Datasets and hyperparameter names and values are represented as standard
       Python, NumPy or SciPy classes rather than new ones whenever feasible.
       This keeps \sklearn easy to use and easy to combine with other libraries.
       % XXX do we need to elaborate more on this, or do we expect everyone
       % to be an experienced Java/adapter pattern hater?
  \item[Composition.]
       Many machine learning tasks are expressible
       as sequences or combinations of transformations to data.
       Some learning algorithms are also naturally viewed
       as meta-algorithms parameterized on other algorithms.
       Whenever possible, those algorithms are implemented and composed from
       existing building blocks.
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       a default and appropriate value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

Whenever practicable, algorithms implemented in \sklearn are thus
directly written in pure Python. For intensive numerical computations though,
the library mostly relies on NumPy vector operations, which allows for the code
to remain concise, readable and efficient. For all algorithms that cannot be
easily and efficiently expressed as NumPy operations, we rely on Cython
\citep{behnel2011cython}  for writing the critical parts and achieving decent
performance and scalability. Cython is a compiled programming language that
extends Python with static typing. It produces efficient C extension modules
that are directly importable from the Python runtime system. Examples of
algorithms written in Cython include stochastic gradient descent for linear
models, some graph-based clustering algorithms or decision trees.  A problem
arising from the use of Cython, however, is that passing pure Python functions
as argument, while possible, degrades performance significantly due to the
runtime checks needed for each function invocation. This is especially
problematic for example for kernel methods, which would benefit from accepting a
kernel function as argument. A potential solution to this problem would be for
\sklearn to provide a Cython-level API, thereby alllowing the user to pass
Cython-implemented kernel functions.

@@

Finally, some estimators in \sklearn also support multi-core architectures.
Machine learning algorithms can indeed often be decomposed into independent
smaller tasks that may each be solved in parallel, thereby reducing the overall
computing times. At this point in time, support for multi-core programming in
Python -- at least in the CPython implementation targeted by scikit-learn -- is
however suboptimal. The interpreter supports multithreading, but due to the
\textit{global interpreter lock}, no two threads in a process may execute Python
code at the same time. It follows that multithreading is not a viable solution
for parallelizing parts of an algorithm, even for embarrassingly parallel
problems, such as fitting and evaluating multiple models in a grid search
procedure. The workaround that we use is to rely on the Python multiprocessing
module, which emulates threads using multiple operating system- level processes.
However this solution is not perfect as it may in turn lead to a large memory
overhead since objects (e.g., input datasets) need to be copied within each
process own memory space. In some cases, it may unfortunately even happen that
the overhead becomes so significant that fully exploiting a multi- core
architecture becomes infeasible due memory constraints. As such, a proper
implementation of parallel machine learning algorithms remain a challenging task
for the project.

>>>>

From its core, the library has been designed to tie in with the set of numeric
and scientific packages centered around the NumPy and SciPy libraries. NumPy
\citep{vanderwalt2011} augments Python with a numeric array data type and proper
array computing primitives while SciPy \citep{varoquaux2013scipy} extends it
further by making common numerical operations available, either by implementing
those in Python/NumPy or by wrapping existing C/C{}\verb!++!/Fortran
implementations. In a similar vein, \sklearn extends Python with a set of
composable machine learning operations.

@@

\begin{description}
  \item[Consistency.]
       All objects (basic or composed) share a consistent interface composed of
       a limited set of methods. This interface is documented in a consistent
       manner for all objects.
  \item[Inspection.]
       Constructor parameters and parameter values determined by learning
       algorithms are stored and exposed as public attributes.
  \item[Non-proliferation of classes.]
       Learning algorithms are the only objects to be of custom classes.
       Datasets and hyper-parameter names and values are represented as standard
       Python, NumPy or SciPy classes rather than new ones whenever feasible.
       This keeps \sklearn easy to use and easy to combine with other libraries.
       % XXX do we need to elaborate more on this, or do we expect everyone
       % to be an experienced Java/adapter pattern hater?
  \item[Composition.]
       Many machine learning tasks are expressible
       as sequences or combinations of transformations to data.
       Some learning algorithms are also naturally viewed
       as meta-algorithms parametrized on other algorithms.
       Whenever possible, those algorithms are implemented and composed from
       existing building blocks.
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       a default and appropriate value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

Whenever practicable, algorithms implemented in \sklearn are thus
directly written in pure Python. For intensive numerical computations though,
the library mostly relies on NumPy vector operations, which allows for the code
to remain concise, readable and efficient. For all algorithms that cannot be
easily and efficiently expressed as NumPy operations, we rely on Cython
\citep{behnel2011cython}  for writing the critical parts and achieving decent
performance and scalability. Cython is a compiled programming language that
extends Python with static typing. It produces efficient C extension modules
that are directly importable from the Python run-time system. Examples of
algorithms written in Cython include stochastic gradient descent for linear
models, some graph-based clustering algorithms or decision trees.  A problem
arising from the use of Cython, however, is that passing pure Python functions
as argument, while possible, degrades performance significantly due to the
run-time checks needed for each function invocation. This is especially
problematic for example for kernel methods, which would benefit from accepting a
kernel function as argument. A potential solution to this problem would be for
\sklearn to provide a Cython-level API, thereby allowing the user to pass
Cython-implemented kernel functions.

@@

Finally, some estimators in \sklearn also support multi-core architectures.
Machine learning algorithms can indeed often be decomposed into independent
smaller tasks that may each be solved in parallel, thereby reducing the overall
computing times. At this point in time, support for multi-core programming in
Python -- at least in the CPython implementation targeted by scikit-learn -- is
however suboptimal. The interpreter supports multithreading, but due to the
\textit{global interpreter lock}, no two threads in a process may execute Python
code at the same time. It follows that multithreading is not a viable solution
for parallelizing parts of an algorithm, even for embarrassingly parallel
problems, such as fitting and evaluating multiple models in a grid search
procedure. The workaround that we use is to rely on the Python multiprocessing
module, which emulates threads using multiple operating system- level processes.
However this solution is not perfect as it may in turn lead to a large memory
overhead since objects (e.g., input datasets) need to be copied within each
process own memory space. In some cases, it may unfortunately even happen that
the overhead becomes so significant that fully exploiting a multi-core
architecture becomes infeasible due memory constraints. As such, a proper
implementation of parallel machine learning algorithms remain a challenging task
for the project.
